---
title: "Differential Methylation"
author: "Lucas Rizkalla"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cerulean
    toc: yes
  pdf_document:
    toc: no
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Set up the environment
library(knitr)
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 10, fig.height = 6) #out.width=700, 
library(pander)
panderOptions('table.split.table', Inf)
set.seed(1)
library(dplyr)
options(stringsAsFactors = FALSE)
```

```{r libraries}
library (RnBeads)
library (writexl)
library (openxlsx)
```

```{r settings}
# Lucas's path
data.dir           = "/Users/lucasrizkalla/Desktop/Methylation Research-VCU/Methylation850K/data"
# Mikhail's path
# data.dir           = "/Users/mdozmorov/Documents/Data/VCU_work/Lathika"
runImport   = TRUE
runDiffMeth = TRUE
idat.dir           = file.path(data.dir, "idats")
sample.annot.abrev = file.path(data.dir, "Sample_Annot_Filt.csv" )
fileNameOut1       = file.path(data.dir, "Differential_Methylation.xlsx")
fileNameOut2       = file.path(data.dir, "RnBeads_Enrichment.xlsx")
```

# Data Import, QC, Preprocessing, and Norm

```{r}
filename = file.path(data.dir, "preprocessed_data.RData")
  
if (runImport) {
  logger.start(fname=NA)
  data.source <- c(idat.dir, sample.annot.abrev)
  rnb.set <- rnb.execute.import(data.source=data.source,
                                data.type="idat.dir")
  
  report.dir = file.path(data.dir, "Report")
  unlink(report.dir, recursive = TRUE) # Delete previous report, if exists, otherwise error
  rnb.run.qc(rnb.set, report.dir)
  
  rnb.set.unfiltered <- rnb.set
  result <- rnb.run.preprocessing(rnb.set, dir.reports=report.dir)
  rnb.set.norm <- result$rnb.set
  save(list = c("rnb.set.norm"), file = filename)
} else {
  load(file = filename)
}
```

# Pairwise Analysis

```{r}
rnb.options("differential.comparison.columns"=c("Visit"),"columns.pairing"=c("Visit"="ID"))
rnb.options("covariate.adjustment.columns"=c("Race", "Age"))
filename1 = file.path(data.dir, "diffMeth_data.RData")

if (runDiffMeth) {
  diffMeth <- rnb.execute.computeDiffMeth(rnb.set.norm, pheno.cols = "Visit")
  comparison <- get.comparisons(diffMeth)[1]
  
  save(list = c("diffMeth"), file = filename1)
} else {
  load(file = filename1)
}
```

##Plotting Differential Methylation Comparison at the Site Level

### Volcano Plot for Differential Methylation (sites)
```{r}
diff.meth.sites <- get.table(diffMeth, comparison, "sites", return.data.frame = TRUE)
#quartz()
plot(diff.meth.sites[, "mean.diff"], -log10(diff.meth.sites[, "diffmeth.p.val"]), xlab="mean difference", ylab="-log10(diffmeth p-value)", col="blue")
```

###Scatterplot for Differential Methylation (sites) with FDR adjusted p-value <0.05
```{r}
isDMS <- diff.meth.sites[,"diffmeth.p.adj.fdr"] < 0.05
#quartz()
create.densityScatter(diff.meth.sites, is.special=isDMS, sparse.points=0.001, add.text.cor=TRUE) + labs(x="mean.beta.A", y="mean.beta.B") + coord_fixed()
```

##Plotting Differential Methylation Comparison at the Region Level

###Volcano Plot for Differential Methylation (promoters)
```{r}
diff.meth.proms <- get.table(diffMeth, comparison, "promoters", return.data.frame = TRUE)
#quartz()
plot(diff.meth.proms[, "mean.mean.diff"], -log10(diff.meth.proms[, "comb.p.val"]), xlab="mean difference", ylab="-log10(combined p-value)", col="blue")
```

###Scatterplot for Differential Methylation (promoters) with Combined Rank among 100 best 
```{r}
diff.meth.proms.ordered <- diff.meth.proms[order(diff.meth.proms$combinedRank, decreasing = FALSE), ]
combinedRank100     <- diff.meth.proms.ordered$combinedRank[100]
combinedRank500     <- diff.meth.proms.ordered$combinedRank[500]

isDMP <- diff.meth.proms.ordered[,"combinedRank"] <= combinedRank100
#quartz()
create.densityScatter(diff.meth.proms, is.special=isDMP, sparse.points=0.001, add.text.cor=TRUE) + labs(x="mean.mean.beta.A", y="mean.mean.beta.B") + coord_fixed()
```

###Scatterplot for Differential Methylation (promoters) with Combined Rank among 500 best 
```{r}
isDMP <- diff.meth.proms.ordered[,"combinedRank"] <= combinedRank500
#quartz()
create.densityScatter(diff.meth.proms, is.special=isDMP, sparse.points=0.001, add.text.cor=TRUE) + labs(x="mean.mean.beta.A", y="mean.mean.beta.B") + coord_fixed()
```

###Scatterplot for Differential Methylation (promoters) with FDR adjusted p-value <0.05 
```{r}
isDMP <- diff.meth.proms[,"comb.p.adj.fdr"] < 0.05
#quartz()
create.densityScatter(diff.meth.proms, is.special=isDMP, sparse.points=0.001, add.text.cor=TRUE) + labs(x="mean.mean.beta.A", y="mean.mean.beta.B") + coord_fixed()
```

###Volcano Plot for Differential Methylation (genes)
```{r}
diff.meth.genes <- get.table(diffMeth, comparison, "genes", return.data.frame = TRUE)
#quartz()
plot(diff.meth.genes[, "mean.mean.diff"], -log10(diff.meth.genes[, "comb.p.val"]), xlab="mean difference", ylab="-log10(combined p-value)", col="blue")
```

###Scatterplot for Differential Methylation (genes) with Combined Rank among 100 best 
```{r}
diff.meth.genes.ordered <- diff.meth.genes[order(diff.meth.genes$combinedRank, decreasing = FALSE), ]
combinedRank100     <- diff.meth.genes.ordered$combinedRank[100]
combinedRank500     <- diff.meth.genes.ordered$combinedRank[500]

isDMG <- diff.meth.genes.ordered[,"combinedRank"] <= combinedRank100
quartz()
#create.densityScatter(diff.meth.genes, is.special=isDMG, sparse.points=0.001, add.text.cor=TRUE) + labs(x="mean.mean.beta.A", y="mean.mean.beta.B") + coord_fixed()

```

###Scatterplot for Differential Methylation (genes) with Combined Rank among 500 best 
```{r}
isDMG <- diff.meth.genes.ordered[,"combinedRank"] <= combinedRank500
#quartz()
create.densityScatter(diff.meth.genes, is.special=isDMG, sparse.points=0.001, add.text.cor=TRUE) + labs(x="mean.mean.beta.A", y="mean.mean.beta.B") + coord_fixed()
```

###Scatterplot for Differential Methylation (promoters) with FDR adjusted p-value <0.05 
```{r}
isDMG <- diff.meth.genes[,"comb.p.adj.fdr"] < 0.05
#quartz()
create.densityScatter(diff.meth.genes, is.special=isDMG, sparse.points=0.001, add.text.cor=TRUE) + labs(x="mean.mean.beta.A", y="mean.mean.beta.B") + coord_fixed()
```

###Volcano Plot for Differential Methylation (CpGIslands)
```{r}
diff.meth.cpgs <- get.table(diffMeth, comparison, "cpgislands", return.data.frame = TRUE)
#quartz()
plot(diff.meth.cpgs[, "mean.mean.diff"], -log10(diff.meth.cpgs[, "comb.p.val"]), xlab="mean difference", ylab="-log10(combined p-value)", col="blue")
```

###Scatterplot for Differential Methylation (CpGIslands) with Combined Rank among 100 best 
```{r}
diff.meth.cpgs.ordered <- diff.meth.cpgs[order(diff.meth.cpgs$combinedRank, decreasing = FALSE), ]
combinedRank100     <- diff.meth.cpgs.ordered$combinedRank[100]
combinedRank500     <- diff.meth.cpgs.ordered$combinedRank[500]

isDMC <- diff.meth.cpgs.ordered[,"combinedRank"] <= combinedRank100
#quartz()
create.densityScatter(diff.meth.cpgs, is.special=isDMC, sparse.points=0.001, add.text.cor=TRUE) + labs(x="mean.mean.beta.A", y="mean.mean.beta.B") + coord_fixed()
```

###Scatterplot for Differential Methylation (CpGIslands) with Combined Rank among 500 best 
```{r}
isDMC <- diff.meth.cpgs.ordered[,"combinedRank"] <= combinedRank500
#quartz()
create.densityScatter(diff.meth.cpgs, is.special=isDMC, sparse.points=0.001, add.text.cor=TRUE) + labs(x="mean.mean.beta.A", y="mean.mean.beta.B") + coord_fixed()
```

###Scatterplot for Differential Methylation (CpGIslands) with FDR adjusted p-value <0.05 
```{r}
isDMC <- diff.meth.cpgs[,"comb.p.adj.fdr"] < 0.05
#quartz()
create.densityScatter(diff.meth.cpgs, is.special=isDMC, sparse.points=0.001, add.text.cor=TRUE) + labs(x="mean.mean.beta.A", y="mean.mean.beta.B") + coord_fixed()
```

#Output of Annotated Results
```{r}
sites.annot     <- annotation(rnb.set.norm, type="sites")
promoter.annot  <- annotation(rnb.set.norm, type="promoters")
genes.annot     <- annotation(rnb.set.norm, type ="genes")
cpgisland.annot <- annotation(rnb.set.norm, type="cpgislands")

annotated.diff.meth.sites <- data.frame(sites.annot, diff.meth.sites)
annotated.diff.meth.proms <- data.frame(promoter.annot, diff.meth.proms)
annotated.diff.meth.genes <- data.frame(genes.annot, diff.meth.genes)
annotated.diff.meth.cpgs  <- data.frame(cpgisland.annot, diff.meth.cpgs)

nrow(annotated.diff.meth.sites)
nrow(annotated.diff.meth.proms)
nrow(annotated.diff.meth.cpgs)
nrow(annotated.diff.meth.genes)

# Save all results
# Note the full results make ~200Mb xlsx
x <- list(annotated.diff.meth.sites, annotated.diff.meth.cpgs, annotated.diff.meth.proms, annotated.diff.meth.genes)
names(x) <- c("Sites", "Cpgs", "Promoters", "Genes")
write.xlsx(x, file = fileNameOut1, row.names = TRUE)
```

#Enrichment Analysis

```{r}
rnb.options("differential.enrichment.go"=TRUE)
rnb.options("differential.enrichment.lola"=TRUE)
# ??? Info message to look into: Not annotated with entrezID --> Skipped
enrich.go <- performGoEnrichment.diffMeth(rnb.set.norm, diffMeth, verbose = TRUE)

enrich.bp.genes.hypo <- enrich.go[["region"]][[comparison]][["BP"]][["genes"]][["rankCut_autoSelect"]][["hypo"]]
enrich.bp.genes.hyper <- enrich.go[["region"]][[comparison]][["BP"]][["genes"]][["rankCut_autoSelect"]][["hyper"]]

enrich.bp.promoter.hypo <- enrich.go[["region"]][[comparison]][["BP"]][["promoters"]][["rankCut_autoSelect"]][["hypo"]]
enrich.bp.promoter.hyper <- enrich.go[["region"]][[comparison]][["BP"]][["promoters"]][["rankCut_autoSelect"]][["hyper"]]

enrich.mf.genes.hypo <- enrich.go[["region"]][[comparison]][["MF"]][["genes"]][["rankCut_autoSelect"]][["hypo"]]
enrich.mf.genes.hyper <- enrich.go[["region"]][[comparison]][["MF"]][["genes"]][["rankCut_autoSelect"]][["hyper"]]

enrich.mf.promoter.hypo <- enrich.go[["region"]][[comparison]][["MF"]][["promoters"]][["rankCut_autoSelect"]][["hypo"]]
enrich.mf.promoter.hyper <- enrich.go[["region"]][[comparison]][["MF"]][["promoters"]][["rankCut_autoSelect"]][["hyper"]]

enrich.bp.genes.hypo     <- data.frame(summary(enrich.bp.genes.hypo))
enrich.bp.genes.hyper    <- data.frame(summary(enrich.bp.genes.hyper))
enrich.bp.promoter.hypo  <- data.frame(summary(enrich.bp.promoter.hypo))
enrich.bp.promoter.hypo  <- data.frame(summary(enrich.bp.promoter.hypo))
enrich.mf.genes.hypo     <- data.frame(summary(enrich.mf.genes.hypo))
enrich.mf.genes.hyper    <- data.frame(summary(enrich.mf.genes.hyper))
enrich.mf.promoter.hypo  <- data.frame(summary(enrich.mf.promoter.hypo))
enrich.mf.promoter.hyper <- data.frame(summary(enrich.mf.promoter.hyper))

y <- list(enrich.bp.genes.hypo, enrich.bp.genes.hyper, enrich.bp.promoter.hypo, enrich.bp.promoter.hypo, enrich.mf.genes.hypo, enrich.mf.genes.hyper, enrich.mf.promoter.hypo, enrich.mf.promoter.hyper)
names(y) <- c("BP Genes Hypo", "BP Genes Hyper", "BP Promoter Hypo", "BP Promoter Hyper", "MF Genes Hypo", "MF Genes Hyper", "MF Promoter Hypo", "MF Promoter Hyper")
write.xlsx(y, file = fileNameOut2, row.names = TRUE)

#lolaDest <- tempfile()
#dir.create(lolaDest)
#lolaDirs <- downloadLolaDbs(lolaDest, dbs="LOLACore")
#lolaDirs[["hg19"]]
#lola <- performLolaEnrichment.diffMeth(rnb.set.norm, diffMeth, lolaDirs[["hg19"]])
# plot
#lolaBarPlot(res$lolaDb, lolaRes, scoreCol="oddsRatio", orderCol="maxRnk", pvalCut=0.05)
```


```{r session_info}
xfun::session_info()
```