---
title: "Analysis"
author: "Mikhail Dozmorov"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: no
  html_document:
    theme: cerulean
    toc: yes
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Set up the environment
library(knitr)
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 10, fig.height = 6) #out.width=700, 
library(pander)
panderOptions('table.split.table', Inf)
set.seed(1)
library(dplyr)
options(stringsAsFactors = FALSE)
```

```{r libraries}
library(readxl)
```

```{r settings}
# Mikhail's path
data.dir           = "/Users/mdozmorov/Documents/Data/VCU_work/Lathika"
fileNameIn1        = file.path(data.dir, "Differential_Methylation.xlsx")

up_dn_separate <- FALSE # Whether to run KEGG separately on up- and downregulated genes. FALSE - do not distinguish directionality
# Which pathway enrichment analysis to run
run_gsea <- FALSE # If TRUE, GSEA pathway enrichment analysis is run, otherwise, standard hypergeometric-based enrichment
```

The `Differential_Methylation.xlsx` file has the results generated with `02_DiffMeth.Rmd`. Each worksheet corresponds to different types of differential analysis, listed in the order of usefulness. The adjusted p-value was set to 0.01.

- "DEGs" - Full results of differential gene analysis, includes coding and noncoding transcripts
    - "DEGs.coding" - Only protein-coding genes
        - "DEGs.coding.up" - Upregulated protein-coding genes
        - "DEGs.coding.dn" - Downregulated protein-coding genes

- "Promoters" - Full results of promoter-centered differential gene analysis, includes coding and noncoding transcripts
    - "Promoters.coding" - Only protein-coding genes
        - "Promoters.coding.up" - Upregulated protein-coding genes
        - "Promoters.coding.dn" - Downregulated protein-coding genes

- "CpGs" - Full results of differential analysis on CpG probe level

- "Islands" - Full results of differential analysis on CpG island level



```{r}

```

