---
title: "Analysis"
author: "Mikhail Dozmorov"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: no
  html_document:
    theme: cerulean
    toc: yes
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Set up the environment
library(knitr)
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 10, fig.height = 6) #out.width=700, 
library(pander)
panderOptions('table.split.table', Inf)
set.seed(1)
library(dplyr)
options(stringsAsFactors = FALSE)
```

```{r}
library(readr)
library(readxl)
library(IlluminaHumanMethylationEPICanno.ilm10b2.hg19)
library(ggplot2)
library(reshape2)
```

```{r}
# Mikhail's path
data.dir           = "/Users/mdozmorov/Documents/Data/VCU_work/Lathika"
fileNameOut2       = file.path(data.dir, "Methylation_bvalues.xlsx")
```

```{r}
# Read methylation data
mtx_expr <- read_xlsx(fileNameOut2, sheet = "bvalues") %>% as.data.frame()
mtx_expr[1:5, 1:5]
mtx_annot <- read_xlsx(fileNameOut2, sheet = "annotations") %>% as.data.frame()
# Read in 850K annotation
data("IlluminaHumanMethylationEPICanno.ilm10b2.hg19")
data("Other")
annot.cg <- data.frame(CpG = rownames(Other), Other)
# Subset to common CpGs
common.cg <- intersect(mtx_expr$CpG, annot.cg$CpG)
mtx_expr <- mtx_expr[mtx_expr$CpG %in% common.cg, ]
annot.cg <- annot.cg[annot.cg$CpG %in% common.cg, ]
annot.cg <- annot.cg[match(mtx_expr$CpG, annot.cg$CpG), ]
all.equal(mtx_expr$CpG, annot.cg$CpG)
```

# CpG

## cpg_selected <- "cg01860693" # mean.diff =  0.20

```{r}
cpg_selected <- "cg01860693" # mean.diff =  0.20
sum(mtx_expr$CpG == cpg_selected)
mtx_to_plot <- data.frame(CpG = mtx_expr[mtx_expr$CpG == cpg_selected, !(colnames(mtx_expr) == "CpG")] %>% as.numeric(), Visit = mtx_annot$Visit)

ggplot(mtx_to_plot, aes(x = Visit, y = CpG)) + geom_boxplot()
```

## cpg_selected <- "cg17798901" # mean.diff = -0.21

```{r}
cpg_selected <- "cg17798901" # mean.diff = -0.21
sum(mtx_expr$CpG == cpg_selected)
mtx_to_plot <- data.frame(CpG = mtx_expr[mtx_expr$CpG == cpg_selected, !(colnames(mtx_expr) == "CpG")] %>% as.numeric(), Visit = mtx_annot$Visit)

ggplot(mtx_to_plot, aes(x = Visit, y = CpG)) + geom_boxplot()
```

# Gene

## gene_selected <- "OR4D9" # mean.mean.diff =  0.10

```{r}
gene_selected <- "OR4D9" # mean.mean.diff =  0.10
sum(grepl(gene_selected, annot.cg$UCSC_RefGene_Name))
mtx_to_plot <- mtx_expr[mtx_expr$CpG %in% annot.cg$CpG[grepl(gene_selected, annot.cg$UCSC_RefGene_Name)], , drop = FALSE] %>% t # Get all CpG for the gene and transpose
colnames(mtx_to_plot) <- mtx_to_plot[1, ] # Make CpG IDs as colnames
mtx_to_plot <- mtx_to_plot[2:nrow(mtx_to_plot), ] # Remove row with colnames
mtx_to_plot <- data.frame(mtx_to_plot, Visit = mtx_annot$Visit) # Add visit column
mtx_to_plot <- melt(mtx_to_plot, id.vars = "Visit") # Long format
colnames(mtx_to_plot) <- c("Visit", "CpG", "Methylation")
mtx_to_plot$Methylation <- as.numeric(mtx_to_plot$Methylation)

ggplot(mtx_to_plot, aes(x = Visit, y = Methylation)) + geom_boxplot() + facet_grid(. ~ CpG)
```

## gene_selected <- "FASLG" # mean.mean.diff = -0.10

```{r}
gene_selected <- "FASLG" # mean.mean.diff = -0.10
sum(grepl(gene_selected, annot.cg$UCSC_RefGene_Name))
mtx_to_plot <- mtx_expr[mtx_expr$CpG %in% annot.cg$CpG[grepl(gene_selected, annot.cg$UCSC_RefGene_Name)], , drop = FALSE] %>% t # Get all CpG for the gene and transpose
colnames(mtx_to_plot) <- mtx_to_plot[1, ] # Make CpG IDs as colnames
mtx_to_plot <- mtx_to_plot[2:nrow(mtx_to_plot), ] # Remove row with colnames
mtx_to_plot <- data.frame(mtx_to_plot, Visit = mtx_annot$Visit) # Add visit column
mtx_to_plot <- melt(mtx_to_plot, id.vars = "Visit") # Long format
colnames(mtx_to_plot) <- c("Visit", "CpG", "Methylation")
mtx_to_plot$Methylation <- as.numeric(mtx_to_plot$Methylation)

ggplot(mtx_to_plot, aes(x = Visit, y = Methylation)) + geom_boxplot() + facet_grid(. ~ CpG)
```


